Quality Assessment of OpenStreetMap Footpath Data

### 2020-06-24 report/report.tex
Spell check

\title{Quality Assesment of OpenStreetMap Footpath Data}
\alttitle{Kvalitetiv undersökning av OpenStreetMaps gångvägsdata}
\author{Alexander Hjelm}
\email{aljhelm@kth.se}
\supervisor{Christopher Peters}
\examiner{Jonas Beskow}
%\hostcompany{Företaget AB} % Remove this line if the project was not done at a host company
\programme{Master in Computer Science}
\school{School of Electrical Engineering and Computer Science}
\date{\today}

@@

\begin{abstract}
This project assesses the feasibility of reconstructing virtual cities from OpenStreetMap (OSM) road and building data, and will focus on the feasibility of including pedestrian footpaths.
It will be examined how common it is that footpaths overlap with other map features such as roads and buildings.
Feature overlap is a type of geomerty error that arises to inaccuracies in the source data, which can lead to strange-looking mesh intersections and artifacts in the final city model.
In this study, a slice of the OSM dataset in the Stockholm area has been taken, and a geometrical algorithm was applied to all road pairs to obtain the average amount of road overlap.
It was found that footpaths were less problematic than car roads in terms of feature overlap, and thus footpaths should pose no more problems in 3D city reproduction than car roads.

As a secondary study, the positional accuracy of the OSM dataset was calculated, since the positional accuracy determines the maximum tolerance in distance between points before it is certain that a road overlaps with another road or building.
Any geodataset has a limit on its positional accuracy, which is determined by the theoretical precision of the measuring methods.
In this project, the accuracy was obtained by comparing coordinate points of OSM building footprints to those in the Swedish University of Agricultural Sciences (SLU) building datset, whose accuracy is known.

@@

Virtual cities has since long been a popular research topic in the field of computer visualization.
The idea of recreating real world cities in 3D applications has a growing number of applications, including visualization and modelling, urban planning and architecture, as well as media and entertainment in the form of video games and 3D animation.
Traditionally, modelling virtual cities is a labour-intensive and time-consuming task due to the need for manual modeling, texturing and design and placement of individual models in the large 3D scene.
This type of tasks is commonly done by professional 3D graphical artists and designers.
However, in recent years there has been major research work in the field of procedural content generation, meaning automatic or semi automatic generation of 3D models without the direct involvment of a 3D artist.
\textbf{Procedural models} typically involve a machine which algorithmically composes together or alters human-generated 3D assets.
Sometimes a degree of computer-generated randomness is induced into the generative algoritm in order to create a larger and less predictable generative space of 3D models that can be created from the same algorithm.

Various procedural methods are indeed used in the field of virtual city reproduction to reduce the labour required to create accurate 3D representations of cities.
These models rely on the availablity of spatial \textbf{geodata}, which serves as the initial input to the procedure of model generation and placement.
Geodata can contain any form of map features, from terrain to road networks and buildings, but should ideally be easily retrievable in a format that the procedural model can easily work with, e.g. as a list of polygons with points as world coordinates.

The field of procedual city reproduction comes with its own host of challenges however.
In practise, urban environments are typically dense and large, spanning a few to hundreds of square kilometers, and are influenced by variables that are hard to quantify, including land policies, government plans, population changes and transportation infrastructure, making it non-trivial to produce 3D city representations with both high accuracy and high detail, while simultaneuosly having high utility in the intended field of study.
\textbf{(Vanegas et al, 2010)}.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio]{img_virtual_city}
    \caption{Real-world cities are complex collections of buildings, blocks and road networks that interconnect them all. Creating 3D virtual cities using methods of procedual generation is an emerging research field that comes with its own set of challenges. This project deals with generating virtual cities from OpenStreetMap data, and the technical challenges therein.$^0$}
    \label{fig:virtual-city}
\end{figure}

@@

This project will strive to determine the feasibility of procedurally reconstructing a city road network from OSM data by using such models as have been mentioned in this section, focusing on the feasibility of including pedestrian footpaths.
\textbf{Footpaths} include any form of path that is designated for pedestrians, including sidewalks, walkways and trails.
OSM commonly divides city road networks into any of the following 4 categories: arterial (primary) roads, secondary roads, residential roads and footpaths.
For the purpose of this report, a \textbf{road} will be any road whose primary use is human travel, either via a vehicle or by foot.
Note that by this definition even sidewalks are classifed as roads, even though they commonly run in connection to larger roads (which are usually reserved for motor vehicles), and thus may not always be thought of as roads in plain language.

In the 3D city model, each road will be represented by a polyline and a fixed \textbf{road width}. The width, or thickness, of a road at a certain point will be defined as the width of the perpendicular road cross section at that point, which is in its entirety usable by the intended types of traffic.
It is important to note that OSM does commonly not include a defition of the width of any road. An optional tag exists that lets users specify the average path width of a road, but this tag is rarely used in practise.

@@

When it comes to the quality of available data, it turns out that in practise, geodata databases without errors are rare.
Geodata validation and correction is only an emerging field of research, and many such datasets were usually obtained years ago, and creating error-free datasets was neither possible due to the lack of validation techniques, nor a high priority.
Luckily there are coming more and more sophisticated methods for automatically correcting the most common errors
\textbf{(Biljecki et al, 2016 (2))}.

Early on, at the stage of importing geodata into a procedural city reproduction model, the most critical source of error is the \textbf{positional accuracy} that the data provider can ensure.
Every geodata service has an upper limit on the positional accuracy of their data, which depends primarily on the method of data acquistion that was used.
Different instruments, such as GPS measurements or satellite image stereoscoping, have theoretical upper limits on their precision, which propagate through the data collection and preservation process as a source of positional error
\textbf{(Fan et al, 2014)}.
Specifically in the case of rendering road networks and buildings, a common issue that stems from low positional accuracy is that the 3D models being generated from road and building data might overlap with each other, which can lead to strange-looking mesh intersections and inaccuracies that modelers often want to avoid.
\textbf{Feature overlap} is a type of geomerty error that arises due to inaccuracies in the source data, often due to limited precision in the measuring methods used to obtain the data.

In this project it was hypothised that the risk of feature overlap may be higher between streets and roads designated for energy efficient mobility, such as footpaths, since they are much more organic and dense in their nature than main streets.
It was also hypothised that one can use the individual road widths and road vertex coordinates as variables when making adjustments against overlap errors.
By first detecting critical locations where feature overlap occurs and subsequently correcting this overlap by adjusting the road width and individual vertex positions within the margin of the positional accuracy, this project examined wether or not it was possible to obtain a slightly altered map with little to no overlap between footpaths and other features.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio]{img_road_error_cases}
    \caption{Example segments of a 3D reconstruciton of the Stockholm road map, with clearly visible geometric errors. The road overlap is marked with red boundaries. The segments show firstly a healthy intersection with main car roads, sidewalks and pedestrian crossings (top left), secondly a city segment where roads have been placed so they overlap with existing buildings due to insufficient data precision (top right), thirdly a similar case where roads overlap with each other (bottom left) and lastly an intersection where the road geomerty twists and self-intersects in strange ways due to invalid source data. This last case will not be examined in this project.}
    \label{fig:img_road_error_cases}
\end{figure}

To limit visual quirks such as mesh self intersection, it is often in the designer's best interest to have a stable procedure for importing road data that results in minimal feature overlap, while still retaining accuracy to the real world.
A possible way of mitigating this problem is looking towards gathering geodata from volunteer-based geoservices such as OSM.
The OSM project is a freely available wiki-like geodatabase based on Volunteered Geographical Information (VGI), meaning anyone can contribute by adding or editing features.
The OSM world dataset is being consistently updated by volonteers from all over the world, and the main idea is that by using OSM geodata for city reproduction, the temporal aspect of data accuracy would in theory be more assured by the crowd contribution nature.
There are already a few such services that use OSM data to generate 3D geodata visualizations for media and video game applications.
A few prominent examples are Mantle
\footnote{Mantle homepage. Accessed 17-05-2020. https://www.mantle.tech/}
and Mapbox
\footnote{Mapbox homepage. Accessed 17-05-2020. https://www.mapbox.com/}.
Such services are however usually limited in the accuracy and type of data that they can provide.
While they can provide both building footprints and road networks, they usually provide road data with no respect to overlap between features.

@@

This project has dealt with the feasibility of including footpath data when generating 3D representations of OSM data, and preserving the integrity of the 3D road network, avoiding overlap and collision between road meshes.
It was hypothised that most road overlap can be corrected by first detecting critical locations where feature overlap occurs, and subsequently adjust the footpath road widths and individual vertex positions within the margin of the OSM positional accuracy, in order to obtain a slightly altered map with little to no overlap between footpaths and other features.

The main research question is as follows:
\begin{center}
    \textbf{Is it technically feasibile to generate a 3D virtual city with footpath road data, with minimal (and preferably no) overlap between 3D features?}
\end{center}

Again, note that feature overlap is defined here as geomerty errors due to inaccuracies in the source data, which results in 3D models overlapping, which can in turn lead to strange-looking mesh intersections.
To investigate the feasibility of generating a 3D virtual city with minimal footpath overlap, the main research question will specifically be divided into two subquestions. The primary subquestion is:

@@

This question aims to find the rate of collision between road features in OSM, and investigate the nature of the most common types of collisions and how they can be corrected.
For the road overlap study, a program has taken a slice of the OSM dataset in the Stockholm area and assigned each road with standard withds according to swedish regulation.
It has then identified critical areas in the OSM dataset, where the paths are so wide that they collide with existing map features.
It has also identified how many of those areas can in theory be corrected by simple translation of the road vertices, without propagation of collision to other features.
Here, features are defined as the geometrical components that comprise the OSM map.
For the purpose of this report, features will refer mainly to roads and building footprints.
For the overlap study, see sections 3.1.2 for a method overview, 3.3 for a detailed walkthrough of the method, and 4.1.1 for the results.

@@

The positional accuracy will be neccessary to carry out the investigation of the first subquestion: the rate of collisions between roads, since it will determine the maximum tolerance in distance between road points before a collision is certain to have occured.
For the secondary research question: the geodata precision study, the positional accuracy will be estimated by comparing feature points between the OSM building dataset and the Swedish University of Agricultural Sciences (SLU) building datset with known accuracy. See section 3.4.2 for an overview of the SLU dataset and how it is obtained.
The positional accuracy will be needed to assess where road overlap occurs and whether the collisions can easily be corrected.
For the positional accuracy study, see sections 3.1.1 for a method overview, 3.2 for a walkthrough of the method, and 4.1.2 for the results.

@@

This study will not consider the lineage aspect of the geodata.
Lineage is defined by \textbf{(Van Oort, 2006)} as the historical aspect of the dataset: how it has been obtained, with what instruments and how it has evolved between the moment it was obtained and the moment it was retrieved.
The entire OSM dataset is under constant modification by volonteers, so it is expected that the accuracy will improve over time. Previous studies highlight how the posiitonal accuracy in similar-sized cities have improved over the course of several years (\textbf{Fan et al, 2014}, \textbf{Haklay, 2010}).
The historical aspect of the OSM dataset is available through the OSM History Viewer
\footnote{OSM History Viewer - OpenStreetMap Wiki. Accessed 20-03-2020. https://wiki.openstreetmap.org/wiki/OSM\_History\_Viewer}
, an online debugging tool that lets anyone freely view the change history of individual fetures in a commit history-like fashion.
There is an option for editors to include a personal note with their changesets to include additional details or motivate why a change was made, but there is no guaratee that the data includes any information on the aquisition method used.

Another aspect of geodata precision is attribute accuracy, which will neither be considered in this study.
Attributes are small flags of metadata that are included with geodata features, letting the user know additional details about a feature that is not reflected in the map, e.g. how many lanes a road has, if the base is asphalt or gravel, or how many stories a building has.
Attribute accuracy is the meausere of how complete attribute information is in a dataset.
How much attribute data is missing and how much of it does not comform with reality?
In general, attribute accuracy is hard to quantify.
In this case however, it might be interesting to compare attributes between the OSM map and the SLU referece map that was used in this project, as both use encoded metadata on a per feature basis, and matching features can be easily identified.
See section 3.4.2 for an overview of the reference dataset that was used in this study.
The attribute accuracy of the two datasets could in theory be assessed by doing a simple feature comparison and creating a translation table between the two attribute maps.
To the knowledge of the author, this has never been done with specifically these two datasets, at the time of writing.

@@

This chapter will dive into the state-of-the-art use of city reconstruction methods within the research fields of visualization and urban design.
This chapter will present the motivations behind digitization of urban planning, as well as a brief history of the research field of OpenStreetMap data qualification, and furthermore present the research papers whose methods are at the foundation of this project.
Lastly, it will present the evaluation metrics that this project served to obtain, and how they relate to the overlap analysis.
For the geodata precision study specifically, this chapter will examine how geodata is qualifyed in general, by examining van Oorts quality criteria, and how these will be used in this project.

@@

In their analysis of prodedural city generation models, \textbf{(Müller et al, 2006)} provide an overview of the process of recreating virtual 3D citites from their real life conterparts.
They present cities as complex collections of \textbf{buildings}, \textbf{blocks}, \textbf{parcels} and \textbf{neighbourhoods} which are all interconnected via a \textbf{street} network, where individual streets can serve different types of traffic.
These features are commonly represented as \textbf{geodata} in map databases, and processed and organized into visualization-ready data by a \textbf{Geographic Information System (GIS)}.
In commercially available geodata, roads are most often defined as polylines: arrays of global coordinate points that come in the same order as they would normally be visited by a person travelling along the road. Buildings in turn are most often defined by their \textbf{building footprints}, a single polygon with global coordinates as vertices, which defines the extents of the building as seen on the map from above.
In practise, cities are often made up out of many square kilometers worth of geodata, and it would be highly labour intensive to manually recreate all these features in 3D.
Instead, \textbf{procedural models} for city reproduction instead rely on programmatic solutions for converting map data to 3D road and building models, and composing them to a 3D world that the designer may continue to fine-tweak through interfaces fo interactive editing.

The \textbf{road network} is what ties the city together as a whole, and forms the grid that divides the lots, quarters and neighbourhoods, giving the city grid its unqigue aestethic character and defining possible building arrangements.
Thus, for any procedural city reproduction model it will be neccessary to have a robust procedure for extracting road geodata and generating the 3D represention of it without compromising the data integrity. 
So far there has been a lot of research into this area, but \textbf{(Stojanovski, 2019)} claims that most of these projects have often focused on the road networks designated for cars and public transit, as a part of the sustainable mobility paradigm and the topic of sustainable cities, particularily due to the interest in self-propelled cars.
The study of energy efficient mobility such as walking and cycling is only in its emerging phase.

@@

This project falls within the broader topic of geodata extraction and modelling, primarily of roads and building parcels.
Geodata modeling is at the foundation of procedural modelling of real-world cities, since any software which models cities based on geodata will need a robust method of aqquiring and laying out road and building data without compromising the data integrity.
In recent years there has been an advent of interactive tools that aid urban planners in placing or generating features of an urban plan, particularly roads and road networks.
Such tools use procedural methods and AI solutions to move the burden of labour from the human designer to the software that the designer uses
\textbf{(Müller et al, 2006)}.

The last decade has seen multiple startups in the area of procedural software for urban arcitecture.
Such software uses procedural models to render a city grid based on real-world locations, or generate new grids and features based on user-specified parameters, and allow for varying degrees of manual editing on grid, road, neighbourhood, city block or individual feature level.
Some of the most prominent examples are ArcGIS Pro (made by Esri), CityEngine (made by the startup Procedural, bought by ESRI) and Urban Canvas (made by the startup Synthicity, bought by Autodesk, now defunct) \textbf{(Stojanovski et al, 2019)}.

Even as such tools are well available, \textbf{(Stojanovski et al, 2020)} note that they are as of today not commonly applied in urbanist practises, due to failures in reflecting the unique workspace of arcitects and their design needs.
Urban arcitecture is a field that not only accounts for spatial data, but also captures many social factors and the individual wills of the many stakeholders.
Each arcitect is an actor in a continously developing environment, and even the archtiect themselves is affecting that environment in a complex manner through their actions.

@@

When assessing the completeness of geodata, a very common scale is Van Oorts criteria for evaluating the quality of geographical information.
At the time of publication, this method of evaluation received much attention from surveyors and cartographers \textbf{(Van Oort, 2006)}, and since then a number of publications have examined slices of geodata by using van Oort's criteria.
In 2008, \textbf{(Haklay, 2010)} conducted such a study to estimate of the quality of OSM positional geodata in London and England.
Since the OSM project started in London it was thought that OSM data in the London metropolitan area would be representative of the highest quality data available, and therefore a good indicator for the whole global OSM dataset.
The study was conducted by comparing OSM data to Ordnance Survey datasets, and it showed that the OSM dataset had a rough positional accuracy of 6 meters from the reference dataset.
However, at the time of the study, the OSM project had only captured roughly 29\% of the area of England, meaning that the data completeness was fairly low, and thus the uncertanity was relatively high.

The next major study to assess the quality of OSM data was published by \textbf{(Kunze, 2012)}.
The study applied different methods to assess the completeness of OSM data in two federal states in Germany, mainly by analysing the area difference between the OSM dataset and an administrative dataset.
Other notable studies about OSM data accuracy were made around the same time in Germany by \textbf{(Zielstra and Zipf, 2010)} and \textbf{(Neis et al, 2012)}, as well as in France by \textbf{(Girres and Touya, 2010)}.

Finally, a major study in 2013-2014, conducted by researchers from various universities in Norway and Germany \textbf{(Fan et al, 2014)}, examined the quality and accuracy in building footprint data of the Munich area, at a time and place where the completeness of the OSM dataset was significantly higher than the study by \textbf{(Haklay, 2010)} (the results of their study showed that the OSM had captured close to 100\% of all building footprints).
This new study included an insight in the geometrical calculations neccessary to match features and points between two datasets, and how to reliably calculate the metrics needed for the most important and assessible of van Oorts criteria.
The results of this study revealed that the positional accuracy of the OSM Munich data at the time was about 4 meters

@@

Programmatic collection of statistics has been neccessary for both the primary and secondary research questions, and this section presents a broad overview of the technical details behind the map analysis and data collection processes for both the geodata precision study and the overlap study.

>>>>

\title{Quality Assessment of OpenStreetMap Footpath Data}
\alttitle{Kvalitativ undersökning av OpenStreetMaps gångvägsdata}
\author{Alexander Hjelm}
\email{aljhelm@kth.se}
\supervisor{Christopher Peters}
\examiner{Jonas Beskow}
%\hostcompany{Företaget AB} % Remove this line if the project was not done at a host company
\programme{Master in Computer Science}
\school{School of Electrical Engineering and Computer Science}
\date{\today}

@@

\begin{abstract}
This project assesses the feasibility of reconstructing virtual cities from OpenStreetMap (OSM) road and building data, and will focus on the feasibility of including pedestrian footpaths.
It will be examined how common it is that footpaths overlap with other map features such as roads and buildings.
Feature overlap is a type of geometry error that arises to inaccuracies in the source data, which can lead to strange-looking mesh intersections and artifacts in the final city model.
In this study, a slice of the OSM dataset in the Stockholm area has been taken, and a geometrical algorithm was applied to all road pairs to obtain the average amount of road overlap.
It was found that footpaths were less problematic than car roads in terms of feature overlap, and thus footpaths should pose no more problems in 3D city reproduction than car roads.

As a secondary study, the positional accuracy of the OSM dataset was calculated, since the positional accuracy determines the maximum tolerance in distance between points before it is certain that a road overlaps with another road or building.
Any geodataset has a limit on its positional accuracy, which is determined by the theoretical precision of the measuring methods.
In this project, the accuracy was obtained by comparing coordinate points of OSM building footprints to those in the Swedish University of Agricultural Sciences (SLU) building dataset, whose accuracy is known.

@@

Virtual cities has since long been a popular research topic in the field of computer visualization.
The idea of recreating real world cities in 3D applications has a growing number of applications, including visualization and modeling, urban planning and architecture, as well as media and entertainment in the form of video games and 3D animation.
Traditionally, modeling virtual cities is a labor-intensive and time-consuming task due to the need for manual modeling, texturing and design and placement of individual models in the large 3D scene.
This type of tasks is commonly done by professional 3D graphical artists and designers.
However, in recent years there has been major research work in the field of procedural content generation, meaning automatic or semi automatic generation of 3D models without the direct involvement of a 3D artist.
\textbf{Procedural models} typically involve a machine which algorithmically composes together or alters human-generated 3D assets.
Sometimes a degree of computer-generated randomness is induced into the generative algoritm in order to create a larger and less predictable generative space of 3D models that can be created from the same algorithm.

Various procedural methods are indeed used in the field of virtual city reproduction to reduce the labor required to create accurate 3D representations of cities.
These models rely on the availability of spatial \textbf{geodata}, which serves as the initial input to the procedure of model generation and placement.
Geodata can contain any form of map features, from terrain to road networks and buildings, but should ideally be easily retrievable in a format that the procedural model can easily work with, e.g. as a list of polygons with points as world coordinates.

The field of procedural city reproduction comes with its own host of challenges however.
In practice, urban environments are typically dense and large, spanning a few to hundreds of square kilometers, and are influenced by variables that are hard to quantify, including land policies, government plans, population changes and transportation infrastructure, making it non-trivial to produce 3D city representations with both high accuracy and high detail, while simultaneously having high utility in the intended field of study.
\textbf{(Vanegas et al, 2010)}.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio]{img_virtual_city}
    \caption{Real-world cities are complex collections of buildings, blocks and road networks that interconnect them all. Creating 3D virtual cities using methods of procedural generation is an emerging research field that comes with its own set of challenges. This project deals with generating virtual cities from OpenStreetMap data, and the technical challenges therein.$^0$}
    \label{fig:virtual-city}
\end{figure}

@@

This project will strive to determine the feasibility of procedurally reconstructing a city road network from OSM data by using such models as have been mentioned in this section, focusing on the feasibility of including pedestrian footpaths.
\textbf{Footpaths} include any form of path that is designated for pedestrians, including sidewalks, walkways and trails.
OSM commonly divides city road networks into any of the following 4 categories: arterial (primary) roads, secondary roads, residential roads and footpaths.
For the purpose of this report, a \textbf{road} will be any road whose primary use is human travel, either via a vehicle or by foot.
Note that by this definition even sidewalks are classified as roads, even though they commonly run in connection to larger roads (which are usually reserved for motor vehicles), and thus may not always be thought of as roads in plain language.

In the 3D city model, each road will be represented by a polyline and a fixed \textbf{road width}. The width, or thickness, of a road at a certain point will be defined as the width of the perpendicular road cross section at that point, which is in its entirety usable by the intended types of traffic.
It is important to note that OSM does commonly not include a definition of the width of any road. An optional tag exists that lets users specify the average path width of a road, but this tag is rarely used in practice.

@@

When it comes to the quality of available data, it turns out that in practice, geodata databases without errors are rare.
Geodata validation and correction is only an emerging field of research, and many such datasets were usually obtained years ago, and creating error-free datasets was neither possible due to the lack of validation techniques, nor a high priority.
Luckily there are coming more and more sophisticated methods for automatically correcting the most common errors
\textbf{(Biljecki et al, 2016 (2))}.

Early on, at the stage of importing geodata into a procedural city reproduction model, the most critical source of error is the \textbf{positional accuracy} that the data provider can ensure.
Every geodata service has an upper limit on the positional accuracy of their data, which depends primarily on the method of data acquisition that was used.
Different instruments, such as GPS measurements or satellite image stereoscoping, have theoretical upper limits on their precision, which propagate through the data collection and preservation process as a source of positional error
\textbf{(Fan et al, 2014)}.
Specifically in the case of rendering road networks and buildings, a common issue that stems from low positional accuracy is that the 3D models being generated from road and building data might overlap with each other, which can lead to strange-looking mesh intersections and inaccuracies that modelers often want to avoid.
\textbf{Feature overlap} is a type of geometry error that arises due to inaccuracies in the source data, often due to limited precision in the measuring methods used to obtain the data.

In this project it was hypothesized that the risk of feature overlap may be higher between streets and roads designated for energy efficient mobility, such as footpaths, since they are much more organic and dense in their nature than main streets.
It was also hypothesized that one can use the individual road widths and road vertex coordinates as variables when making adjustments against overlap errors.
By first detecting critical locations where feature overlap occurs and subsequently correcting this overlap by adjusting the road width and individual vertex positions within the margin of the positional accuracy, this project examined whether or not it was possible to obtain a slightly altered map with little to no overlap between footpaths and other features.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio]{img_road_error_cases}
    \caption{Example segments of a 3D reconstruction of the Stockholm road map, with clearly visible geometric errors. The road overlap is marked with red boundaries. The segments show firstly a healthy intersection with main car roads, sidewalks and pedestrian crossings (top left), secondly a city segment where roads have been placed so they overlap with existing buildings due to insufficient data precision (top right), thirdly a similar case where roads overlap with each other (bottom left) and lastly an intersection where the road geometry twists and self-intersects in strange ways due to invalid source data. This last case will not be examined in this project.}
    \label{fig:img_road_error_cases}
\end{figure}

To limit visual quirks such as mesh self intersection, it is often in the designer's best interest to have a stable procedure for importing road data that results in minimal feature overlap, while still retaining accuracy to the real world.
A possible way of mitigating this problem is looking towards gathering geodata from volunteer-based geoservices such as OSM.
The OSM project is a freely available wiki-like geodatabase based on Volunteered Geographical Information (VGI), meaning anyone can contribute by adding or editing features.
The OSM world dataset is being consistently updated by volunteers from all over the world, and the main idea is that by using OSM geodata for city reproduction, the temporal aspect of data accuracy would in theory be more assured by the crowd contribution nature.
There are already a few such services that use OSM data to generate 3D geodata visualizations for media and video game applications.
A few prominent examples are Mantle
\footnote{Mantle homepage. Accessed 17-05-2020. https://www.mantle.tech/}
and Mapbox
\footnote{Mapbox homepage. Accessed 17-05-2020. https://www.mapbox.com/}.
Such services are however usually limited in the accuracy and type of data that they can provide.
While they can provide both building footprints and road networks, they usually provide road data with no respect to overlap between features.

@@

This project has dealt with the feasibility of including footpath data when generating 3D representations of OSM data, and preserving the integrity of the 3D road network, avoiding overlap and collision between road meshes.
It was  that most road overlap can be corrected by first detecting critical locations where feature overlap occurs, and subsequently adjust the footpath road widths and individual vertex positions within the margin of the OSM positional accuracy, in order to obtain a slightly altered map with little to no overlap between footpaths and other features.

The main research question is as follows:
\begin{center}
    \textbf{Is it technically feasible to generate a 3D virtual city with footpath road data, with minimal (and preferably no) overlap between 3D features?}
\end{center}

Again, note that feature overlap is defined here as geometry errors due to inaccuracies in the source data, which results in 3D models overlapping, which can in turn lead to strange-looking mesh intersections.
To investigate the feasibility of generating a 3D virtual city with minimal footpath overlap, the main research question will specifically be divided into two subquestions. The primary subquestion is:

@@

This question aims to find the rate of collision between road features in OSM, and investigate the nature of the most common types of collisions and how they can be corrected.
For the road overlap study, a program has taken a slice of the OSM dataset in the Stockholm area and assigned each road with standard widths according to swedish regulation.
It has then identified critical areas in the OSM dataset, where the paths are so wide that they collide with existing map features.
It has also identified how many of those areas can in theory be corrected by simple translation of the road vertices, without propagation of collision to other features.
Here, features are defined as the geometrical components that comprise the OSM map.
For the purpose of this report, features will refer mainly to roads and building footprints.
For the overlap study, see sections 3.1.2 for a method overview, 3.3 for a detailed walk through of the method, and 4.1.1 for the results.

@@

The positional accuracy will be necessary to carry out the investigation of the first subquestion: the rate of collisions between roads, since it will determine the maximum tolerance in distance between road points before a collision is certain to have occurred.
For the secondary research question: the geodata precision study, the positional accuracy will be estimated by comparing feature points between the OSM building dataset and the Swedish University of Agricultural Sciences (SLU) building dataset with known accuracy. See section 3.4.2 for an overview of the SLU dataset and how it is obtained.
The positional accuracy will be needed to assess where road overlap occurs and whether the collisions can easily be corrected.
For the positional accuracy study, see sections 3.1.1 for a method overview, 3.2 for a walk through of the method, and 4.1.2 for the results.

@@

This study will not consider the lineage aspect of the geodata.
Lineage is defined by \textbf{(Van Oort, 2006)} as the historical aspect of the dataset: how it has been obtained, with what instruments and how it has evolved between the moment it was obtained and the moment it was retrieved.
The entire OSM dataset is under constant modification by volunteers, so it is expected that the accuracy will improve over time. Previous studies highlight how the positional accuracy in similar-sized cities have improved over the course of several years (\textbf{Fan et al, 2014}, \textbf{Haklay, 2010}).
The historical aspect of the OSM dataset is available through the OSM History Viewer
\footnote{OSM History Viewer - OpenStreetMap Wiki. Accessed 20-03-2020. https://wiki.openstreetmap.org/wiki/OSM\_History\_Viewer}
, an online debugging tool that lets anyone freely view the change history of individual features in a commit history-like fashion.
There is an option for editors to include a personal note with their changesets to include additional details or motivate why a change was made, but there is no guarantee that the data includes any information on the acquisition method used.

Another aspect of geodata precision is attribute accuracy, which will neither be considered in this study.
Attributes are small flags of metadata that are included with geodata features, letting the user know additional details about a feature that is not reflected in the map, e.g. how many lanes a road has, if the base is asphalt or gravel, or how many stories a building has.
Attribute accuracy is the measure of how complete attribute information is in a dataset.
How much attribute data is missing and how much of it does not conform with reality?
In general, attribute accuracy is hard to quantify.
In this case however, it might be interesting to compare attributes between the OSM map and the SLU reference map that was used in this project, as both use encoded metadata on a per feature basis, and matching features can be easily identified.
See section 3.4.2 for an overview of the reference dataset that was used in this study.
The attribute accuracy of the two datasets could in theory be assessed by doing a simple feature comparison and creating a translation table between the two attribute maps.
To the knowledge of the author, this has never been done with specifically these two datasets, at the time of writing.

@@

This chapter will dive into the state-of-the-art use of city reconstruction methods within the research fields of visualization and urban design.
This chapter will present the motivations behind digitization of urban planning, as well as a brief history of the research field of OpenStreetMap data qualification, and furthermore present the research papers whose methods are at the foundation of this project.
Lastly, it will present the evaluation metrics that this project served to obtain, and how they relate to the overlap analysis.
For the geodata precision study specifically, this chapter will examine how geodata is qualified in general, by examining van Oort's quality criteria, and how these will be used in this project.

@@

In their analysis of procedural city generation models, \textbf{(Müller et al, 2006)} provide an overview of the process of recreating virtual 3D cities from their real life counterparts.
They present cities as complex collections of \textbf{buildings}, \textbf{blocks}, \textbf{parcels} and \textbf{neighborhoods} which are all interconnected via a \textbf{street} network, where individual streets can serve different types of traffic.
These features are commonly represented as \textbf{geodata} in map databases, and processed and organized into visualization-ready data by a \textbf{Geographic Information System (GIS)}.
In commercially available geodata, roads are most often defined as polylines: arrays of global coordinate points that come in the same order as they would normally be visited by a person traveling along the road. Buildings in turn are most often defined by their \textbf{building footprints}, a single polygon with global coordinates as vertices, which defines the extents of the building as seen on the map from above.
In practice, cities are often made up out of many square kilometers worth of geodata, and it would be highly labor intensive to manually recreate all these features in 3D.
Instead, \textbf{procedural models} for city reproduction instead rely on programmatic solutions for converting map data to 3D road and building models, and composing them to a 3D world that the designer may continue to fine-tweak through interfaces for interactive editing.

The \textbf{road network} is what ties the city together as a whole, and forms the grid that divides the lots, quarters and neighborhoods, giving the city grid its unique aestethic character and defining possible building arrangements.
Thus, for any procedural city reproduction model it will be necessary to have a robust procedure for extracting road geodata and generating the 3D representation of it without compromising the data integrity. 
So far there has been a lot of research into this area, but \textbf{(Stojanovski, 2019)} claims that most of these projects have often focused on the road networks designated for cars and public transit, as a part of the sustainable mobility paradigm and the topic of sustainable cities, particularly due to the interest in self-propelled cars.
The study of energy efficient mobility such as walking and cycling is only in its emerging phase.

@@

This project falls within the broader topic of geodata extraction and modeling, primarily of roads and building parcels.
Geodata modeling is at the foundation of procedural modeling of real-world cities, since any software which models cities based on geodata will need a robust method of acquiring and laying out road and building data without compromising the data integrity.
In recent years there has been an advent of interactive tools that aid urban planners in placing or generating features of an urban plan, particularly roads and road networks.
Such tools use procedural methods and AI solutions to move the burden of labor from the human designer to the software that the designer uses
\textbf{(Müller et al, 2006)}.

The last decade has seen multiple startups in the area of procedural software for urban architecture.
Such software uses procedural models to render a city grid based on real-world locations, or generate new grids and features based on user-specified parameters, and allow for varying degrees of manual editing on grid, road, neighborhood, city block or individual feature level.
Some of the most prominent examples are ArcGIS Pro (made by Esri), CityEngine (made by the startup Procedural, bought by ESRI) and Urban Canvas (made by the startup Synthicity, bought by Autodesk, now defunct) \textbf{(Stojanovski et al, 2019)}.

Even as such tools are well available, \textbf{(Stojanovski et al, 2020)} note that they are as of today not commonly applied in urbanist practices, due to failures in reflecting the unique workspace of architects and their design needs.
Urban architecture is a field that not only accounts for spatial data, but also captures many social factors and the individual wills of the many stake holders.
Each architect is an actor in a continuously developing environment, and even the architect themselves is affecting that environment in a complex manner through their actions.

@@

When assessing the completeness of geodata, a very common scale is Van Oort's criteria for evaluating the quality of geographical information.
At the time of publication, this method of evaluation received much attention from surveyors and cartographers \textbf{(Van Oort, 2006)}, and since then a number of publications have examined slices of geodata by using van Oort's criteria.
In 2008, \textbf{(Haklay, 2010)} conducted such a study to estimate of the quality of OSM positional geodata in London and England.
Since the OSM project started in London it was thought that OSM data in the London metropolitan area would be representative of the highest quality data available, and therefore a good indicator for the whole global OSM dataset.
The study was conducted by comparing OSM data to Ordnance Survey datasets, and it showed that the OSM dataset had a rough positional accuracy of 6 meters from the reference dataset.
However, at the time of the study, the OSM project had only captured roughly 29\% of the area of England, meaning that the data completeness was fairly low, and thus the uncertainty was relatively high.

The next major study to assess the quality of OSM data was published by \textbf{(Kunze, 2012)}.
The study applied different methods to assess the completeness of OSM data in two federal states in Germany, mainly by analyzing the area difference between the OSM dataset and an administrative dataset.
Other notable studies about OSM data accuracy were made around the same time in Germany by \textbf{(Zielstra and Zipf, 2010)} and \textbf{(Neis et al, 2012)}, as well as in France by \textbf{(Girres and Touya, 2010)}.

Finally, a major study in 2013-2014, conducted by researchers from various universities in Norway and Germany \textbf{(Fan et al, 2014)}, examined the quality and accuracy in building footprint data of the Munich area, at a time and place where the completeness of the OSM dataset was significantly higher than the study by \textbf{(Haklay, 2010)} (the results of their study showed that the OSM had captured close to 100\% of all building footprints).
This new study included an insight in the geometrical calculations necessary to match features and points between two datasets, and how to reliably calculate the metrics needed for the most important and assessable of van Oort's criteria.
The results of this study revealed that the positional accuracy of the OSM Munich data at the time was about 4 meters

@@

Programmatic collection of statistics has been necessary for both the primary and secondary research questions, and this section presents a broad overview of the technical details behind the map analysis and data collection processes for both the geodata precision study and the overlap study.
